---
title: "Eye Health Facilities"
author: "Kevin Tang, John Nesemann, Ian McCormick"
date: "6/30/2021"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
## Set up

Load libraries
```{r}
library(tidyverse)            # for data cleaning and data viz
library(lubridate)            # for dates
library(here)                 # for directory management
library(sf)                   # for spatial data processing
theme_set(theme_minimal())    # set ggplot theme
```

Read data
```{r}
eye_fac <- read.csv(here("data", "Malawi_facilities_public_gps.csv"))
eye_fac_sf <- eye_fac %>% st_as_sf(coords = c("long", "lat"), crs=4326)
eye_fac_sf_t <- st_transform(eye_fac_sf, crs=2163)

malawi <- st_read(here("data", "geo_shp", "mwi_admbnda_adm2_nso_20181016.shp"))
```

## Eye health facilities in Malawi

Eye health facilities are defined as public sector health facilities that have a permanent member of staff who is trained in ophthalmic care. Types of facilities can range from small health clinics to central hospitals. 
```{r}
ggplot() +
  geom_sf(data = malawi) +
  geom_sf(data = eye_fac_sf_t, aes(color = cataract_surgical_services)) +
  theme_void() 
```

## Questions
(Kevin) What eye health outcomes are we interested in and what does the referral system look like in LMICs (i.e., Malawi)? For instance, in this Malawi example, we have a number of health facilities with trained ophthalmic staff but no capacity to perform cataract surgical services. So what kind of care can be expected in these facilities and at what point clinically are patients referred to larger health centers with more robust eye care infrastructure?
(Ian) I think the best bet would be to start by looking at travel time to cataract surgical services (in pop 50+) rather than all public eye health facilities (although the third indicator requires consideration of travel time between them). The primary eye care services available from the non-cataract surgical facilities will be varied and essentially unknowable at scale. Patients with cataract and a certain level of impairment will be referred to secondary care. The visual acuity threshold for referral varies according to service maturity. E.g. in Malawi there will be lots of cataract blind people, so those with lower levels of vision impairment will not be prioritised. Patients will be referred for other non-cataract conditions requiring treatment too. Ophthalmic nurses will prescribe antibiotics, anti-inflammatories, anti-allergy meds as required/available.
Focusing on cataract we can link services to cataract vision impairment and effective cataract surgical coverage (eCSC) outcomes in regions where it is measured in the population and CSR (cataract surgical output – ops per million per year) from facilities.

--

(Kevin) Do surveillance and community case finding programs exist for eye diseases? There are two sides to geographic access to care: (1) individuals with disease who travel to seek care and (2) health systems that expend resources to find cases in communities. It would be good to get an idea about if/what kinds of programs exist for the latter.
(Ian) On case finding: broadly yes, there will be community cataract screening +/- cataract surgical camps. I will ask Shaffi in Malawi for an overview of current practice.
Generally, the trend is to move away from delivering cataract surgery in camps as outcomes may be worse and follow-up care less structured, so even where the health system identifies people as cataract vision impaired in the community those people may well be travelling to access surgery at secondary centres. Or not… depending on their motivation, financial means and physical ability to make the journey.

##Friction Surface

Download Packages
```{r}
## Packages
library(gdistance)
library(sp)
library(abind)
library(rje)
library(ggplot2)
library(malariaAtlas)
## Plot defaults
theme_set(theme_minimal(base_size=14))
```

Clip friction surface for Malawi
```{r}
malawi_clip <- readRDS((here("data", "geo_shp", "gadm36_MWI_0_sp.rds")))
plot(malawi_clip, main="Shape for Clipping")

friction <- malariaAtlas::getRaster(
                      surface = "A global friction surface enumerating land-based travel speed for a nominal year 2015",
                      shp = malawi_clip)
malariaAtlas::autoplot_MAPraster(friction)
```

Convert friction surface to a transition matrix
```{r}
T <- gdistance::transition(friction, function(x) 1/mean(x), 8) 
T.GC <- gdistance::geoCorrection(T)   
```

Point locations (Eye care facilities)
```{r}
## Point locations
point.locations <- eye_fac %>% dplyr::select(long, lat, name, cataract_surgical_services)
names(point.locations) <- c("X_COORD", "Y_COORD", "name", "cataract_surgical_services")

coordinates(point.locations) <- ~ X_COORD + Y_COORD #Keep only point coordinates within the shapefile bounds

points <- as.matrix(point.locations@coords) #convert to matrix of lat-longs for accessibility algorithm
```

Final raster model
```{r}
access.raster <- gdistance::accCost(T.GC, points) #run an “accumulated cost surface” algorithm to calculate travel time to closest facility

p <- malariaAtlas::autoplot_MAPraster(access.raster, 
                        shp_df=malawi_clip, printed=F)

full_plot <- p[[1]] + geom_point(data=data.frame(point.locations@coords), 
                           aes(x=X_COORD, y=Y_COORD)) +
scale_fill_gradientn(colors = rev(rje::cubeHelix(gamma=1.0, 
                                                 start=1.5, 
                                                 r=-1.0, 
                                                 hue=1.5, 
                                                 n=16)), 
                     name="Minutes \n of Travel") + 
ggtitle("Travel Time to Most Accessible Eye Care Facility") +
theme(axis.text=element_blank(),
      panel.border=element_rect(fill=NA, color="white"))

print(full_plot)
```

